---
title: "Assigment MP_6"
author: "Ritik Kumar"
date: "2024-10-18"
output: html_document
---

## Import Data
```{r}
library(fpp)
library(fpp2)
library(TTR)
library(ggplot2)
library(readr)
library(dplyr)
library(forecast)

flightvolume <- read_csv("C:/Users/malho/Downloads/flightVolume.csv")
View(flightvolume)
```
##### Considering only Reslong (long-distance residential flights) column from 1976 to 2022
```{r}
flightvolume_ts <- ts(flightvolume$reslong, start=c(1976, 1), frequency=12)
plot(flightvolume_ts)
```

*We'll assume the data starts from January 1976 and goes monthly
*if we consider the whole data that might not be give us the exact forecast.
*from 1976 to 2022, this data should be good enough to be considered for forecasting.
*Reslong represents long-distance residential flights


## Plot and Inference
##### 1.Creating Times Series Plot
```{r}
plot(flightvolume_ts, main = " Reslong Time Series (1976-2022) Plot ", xlab = 'Year', ylab = "Frequency of flights")
```

* we start with plotting the time series for visualize and understand the data.

##### Initial Oberversation 
* The plot shows that there is an decreasing trend in the frequency of flights from year 1976 till around 1987.
* From 1990 till 2008, there has been a steady increasing trend in the flights frequency.
* From 2010, there has been again drastically decreasing trend till the year 2022.

## 2.Verify how much history to include in your analysis.

```{r}
summary(flightvolume_ts)
plot(flightvolume_ts, main="Reslong Time Series (1976-2022) Plot", ylab="Frequency of flights", xlab="Year")
```

* The summary function above gives the min, max, mean, median, 1st and 3rd Quarterly values of the times series.
* After that we plot the graph to verify how much history to include in our analysis

## 3.Hypothesize if the dataset has trend, seasonality, or both. 
##### 4. Verifying using ACF plot to check for seasonality or autocorrelation
```{r}
Acf(flightvolume_ts, main = " Acf plot for Reslong Time Series")
```

##### 5.Verifying using decomposition
```{r}
stl_decomp <- stl(flightvolume_ts, s.window = "periodic")
plot(stl_decomp, main = 'Decomposition Plot')
```

# Decompose the time series to verify trend and seasonality
```{r}
decomp <- decompose(flightvolume_ts)
plot(decomp)
decomp$type
```

* The decomposition seems to be additive. 
* Trend: Long-term movement in the data
* Seasonal: Repeated cycles, usually yearly
* Residual: Random fluctuations that remain after removing trend and seasonality
* Because, with as trend increases, we do not see any strong increase in the seasonality. The seasonality appears to be the same throughout.
* Reslong time series shows a trend but does not display strong seasonality.

## 6.Choose an accuracy measure
* We'll use Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) for model evaluation. later

## 7.Create a forecast model for the next 12 months. Include Naive, Average, Exponential Smoothing, HoltWinters, and Decomposition (both types). 

## Naive model
```{r}
naive_forecast <- naive(flightvolume_ts, h=12)
plot(naive_forecast)
```

##### Naive Method Summary
*The naive model simply assumes the future will be exactly like the most recent value (no trend or seasonality).

## Simple Moving Averages
```{r}
MA5_forecast <- ma(flightvolume_ts,order=5)
MA9_forecast <- ma(flightvolume_ts,order=9)
plot(flightvolume_ts, main ="Plot along with moving averages")
lines(MA5_forecast,col="Red")
lines(MA9_forecast,col="Blue")
```

##### Observations
* The average model assumes the future will be the same as the historical average.
* From the plots, it is observed that the higher the order we consider, the smoother the moving average curve in the plot.
* It can be seen that the blue line above is the smoothest compared to Red line.
* The Red line (order 5) gives the most real data compared to the other line. The higher order averages smoother the plot and do not give the actual values.

## Exponential Smoothing Model (Simple Exponential Smoothing - SES)
```{r}
ses_forecast <- ses(flightvolume_ts, h=12)
plot(ses_forecast)
attributes(ses_forecast)
```

##### Observations 
* Exponential Smoothing gives more weight to recent observations, making it more responsive to changes.
* From 1990 to 2008 there is observed to be an increasing trend in the data. So, this forecast may not be a right way to forecast.

## Holt-Winters model - Captures both trend and seasonality.
```{r}
hw_model <- hw(flightvolume_ts)
hw_forecast <- forecast(hw_model, h=12)
plot(hw_forecast, main="Holt-Winters Forecast")
attributes(hw_forecast)
```

* Here, The Holt-Winters model considers both trend and seasonality, making it suitable for seasonal data.
* This is because the seasonality isn't increasing with trend. This is an additive time series.

###### Observations
```{r}
hw_model$model
```

* Alpha = 0.3948 Alpha specifies the coefficient for the level smoothing in Holt winters.
* Beta = 1e-04 Beta specifies the coefficient for the trend smoothing in Holt winters. 
* Gamma = 0.4462 Gamma specifies the coefficient for the seasonal smoothing in Holt winters.
* Values 1.0 means that the latest value has highest weight.
* Initial states:
    l = 5.1498 
    b = -7e-04 
    s = 0.3443 -1.2489 -1.3403 -1.3922 0.614 -0.1693
           -0.7754 0.6123 0.6282 0.4555 -0.5577 2.8294
* Sigma = 0.4351 Sigma defines the variance of the forecast values.

## Decomposition-based forecast Model
```{r}
decomp_trend <- tslm(flightvolume_ts ~ trend + season)
decomp_forecast <- forecast(decomp_trend, h=12)
plot(decomp_forecast, main="Decomposition-Based Forecast")
```

## Decomposition (Additive)
##### Perform additive decomposition
```{r}
add_decomp <- decompose(flightvolume_ts, type = "additive")
plot(add_decomp)
```

* Decomposition splits the time series into trend, seasonal, and residual components.

## Decomposition (Multiplicative)
##### Perform multiplicative decomposition
```{r}
mult_decomp <- decompose(flightvolume_ts, type = "multiplicative")
plot(mult_decomp)
```

* Multiplicative decomposition is used when the seasonal component changes proportionally to the trend.

## 8.Show model rank with accuracy measures (MAE and RMSE)
```{r}
naive_acc <- accuracy(naive_forecast)
naive_acc
ses_acc <- accuracy(ses_forecast)
ses_acc
hw_acc <- accuracy(hw_forecast)
hw_acc
decomp_acc <- accuracy(decomp_forecast)
decomp_acc

# Create a summary of accuracy measures (MAE and RMSE)
accuracy_summary <- data.frame(
  Model = c("Naive","Exponential Smoothing", "Holt-Winters", "Decomposition"),
  MAE = c(naive_acc[1, "MAE"], ses_acc[1, "MAE"], hw_acc[1, "MAE"], decomp_acc[1, "MAE"]),
  RMSE = c(naive_acc[1, "RMSE"], ses_acc[1, "RMSE"], hw_acc[1, "RMSE"], decomp_acc[1, "RMSE"])
)

print(accuracy_summary)
```

* This table ranks the models based on their accuracy using MAE and RMSE. Lower values indicate better accuracy.

## 9.Choose which models to use for forecasting

* Based on the accuracy measures, we would select the model with the lowest MAE and RMSE, typically the Holt-Winters or Decomposition model.

##### Best & Worst Forecasts

* To start with, there is nothing like best or worst forecast.
* Considering the accuracy data above, HoltWinters forecast seems to fit the time series the best as it has the least error values.
* And naive forecast seems to be the worst as it has the largest ME and RMSE values.

## 10. Provide the forecast for the next 12 months (point and range)
```{r}
final_forecast <- forecast(hw_model, h=12)
print(final_forecast)
plot(final_forecast, main="Final Forecast for the Next 12 Months")
```

* This is the final forecast, providing point estimates for the next 12 months along with confidence intervals.

## Conclusion 
* The data seemed to have strong trend and weak seasonality initially and we checked the same with Acf and confirmed it.
* Based on the forecasting models like naive, Exponential Smoothing Model, and HoltWinters, we can see that HoltWinters forecast provides to be the better forecasting method in this case.
* This is because the forecast fits perfectly and also the error values are quiet low for HoltWinters forecast.
* This shows that our hypothesis is correct based on the accuracy of all the models.
* Based on the analysis and forecast, the time series will decrease over the next year and the next 2 years.